\begin{comment}
\documentclass{book}
\usepackage{master}
\newcommand{\rec}{$\text{R\'ecoltes et Semailles}$}
\newcommand{\no}{n$^\circ$}
\hfuzz = 100pt
\begin{document}
\end{comment}

\chapter{By way of a foreword}
\label{chapter:1}

\marginpar{p. A1}
January 30 1986

In order for R\'ecoltes et Semailles to be ready to be sent to the publisher, all that was left to write was the foreword.
And I swear that I went into it with all the goodwill in the world to write something suitable. 
Something \textbf{reasonable} this time. No more than three or four carefully phrased pages that would introduce this huge volume of over a thousand
pages. Something which ``grabs'' the attention of the jaded reader, 
perhaps suggesting to him that in these frightening ``over a thousand pages'', he might be able to find
things of interest (or things which concern him, who knows?). It is not
really my style to pander to others. But I was ready to make an exception for once!
The ``publisher crazy enough to give it a shot'' (by being willing to publish this visibly unpublishable
monster) had to make ends meet one way or another. 

But then, it didn't happen. Even though I tried my best. And not only over the course of an afternoon, as I had
originally planned. Tomorrow will mark three weeks since I started writing the foreword, since the sheets began
accumulating.
What came out, to be sure, isn't what one could reasonably call a ``foreword''.
I guess it is yet another miss! Blame it on my old age - I have never been much of a salesman. Even when
it comes to pleasing (oneself or one's friends\ldots).

What came out instead is a sort of long ``Walk'', with commentary, through my life work as a mathematician. 
A Walk intended mostly for the ``layman'' - he who ``never understood anything about
math''; as well as for myself, having never indulged in such a Walk. 
Step by step, I found myself unraveling and writing things that had previously remained
unspoken. As if by chance, these are also the things which I find most essential, both in
my practice and in determining its outcome. 
They are things which are not technical in nature. It will be up to you, the reader, to decide whether
or not I succeeded in my naive enterprise to ``get the message through'' - an enterprise
which is surely a bit mad. 
My satisfaction and pleasure will come from making you feel these things. Many of my wise colleagues no longer feel them, 
possibly because they have become too wise
and too prestigious - a situation which often leads to losing touch with the simplest and most essential things in life. 

During this ``Walk through a life work'' 
I also speak about my life,
as well as, here and there, what 
R\'ecoltes et Semailles as a whole
is about.
I go into this in more detail in the ``Letter'' 
\marginpar{p. A2}
(dated from May of last year) which
follows the ``Walk''. 
This Letter was addressed to my
previous students and to my ``old friends'' in the mathematical community. 
But even the Letter is not technical in nature. 
It can be read by anyone
interested in learning,
through a ``heartfelt'' narrative, about the various reasons that led me to writing 
R\'ecoltes et Semailles. Even more than the Walk, the Letter will provide a 
preview of a particular prevailing atmosphere in the ``prestigious mathematical world'',
as well as a preview of my peculiar writing style (the same could be said of the Walk) and of the
spirit that is expressed in this style - a spirit which is not universally
appreciated. 

In the Walk and throughout R\'ecoltes et Semailles, I speak about the act of 
\textbf{doing mathematics}.
It is an activity for which I have first-hand experience and which I know very well. Most of the
things I say about it can surely be said of any kind of creative work, or of any work involving
discovery - it is at the very least true of ``intellectual''
work, mostly involving ``using one's brain'' and writing. 
All such work proceeds through the formation and deepening of an
\textbf{understanding} of the things which are being probed. 
To take an example at the 
opposite extreme, romantic passion is also an act of discovery. 
It opens us to understanding of a ``physical'' nature which also is also renewed, developed, and deepened over time. 
Both of these impulses - 
that which, say, animates a mathematician at work and
that which enlivens a lover 
- are much closer in nature than we generally assume or will readily admit. 
I hope that the pages of 
R\'ecoltes et Semailles 
will contribute to making you feel this impulse in your work and in your daily life. 

Most of the Walk focuses on mathematical work itself. 
I mostly remain silent on the \textbf{context} in which this work takes place, and
on the \textbf{motivations} at play outside of mathematical work itself. 
This comes at the risk of painting a flattering but deformed image of myself, of mathematicians, or of ``scientists'' in general - think ``grand and noble passion'' without any further rectification.
In accordance, that is, with the great ``Myth of Science" 
(with a capital S, thank you very mich!).
A heroic myth, ``promethean'' in nature, to which countless writers and thinkers have succumbed 
(and continue to succumb). 
Only historians, if that, manage to sometimes resist this tantalizing myth. 
The truth is, among the things which motivate these ``scientists'',
sometimes even inciting them to devote themselves entirely to their work, ambition and vanity
play just as important and universal a role as they do in any other profession. 
This phenomenon 
\marginpar{p. A3}
appears in
more or less subtle ways depending on the person. I am no exception to this pattern -
the reading of my testimony will hopefully settle this fact beyond doubt. 

It is true also that even the most intense of ambition is
powerless when it comes to discovering or proving a novel mathematical statement - just as it is
powerless (for instance) to ``make one hard'' (in the literal sense of the term).
Whether man or woman, what ``makes one hard'' is not ambition, nor the desire to shine or to
display power (of a sexual nature in this case) - quite to the contrary!
It is the acute perception of something strong, at once very real and very delicate. 
One could call it ``beauty'', though this is only one of
this thing's thousand faces. Being ambitious doesn't prevent one from 
sensing the beauty of a being or a thing. But it is \textbf{not} ambition which makes us
feel it\ldots 

The person who first discovered and mastered fire was somebody just like you and me -
not at all the kind of being we refer to as ``hero'', ``demi-god'', and so on.  
Surely, just like you and me, he found himself in the grip of anxiety as well as the 
time-worn fixture of vanity which makes one forget about it.
But at the instant at which he ``knew'' fire, there was neither fear nor vanity. 
Such is the truth lying within the heroic myth.
The myth turns insipid, it becomes a fixture, when it is used
to conceal another aspect of things which is just as real and essential. 

My aim in R\'ecoltes et Semailles has been to address both of the above aspects of the heroic myth - 
the impulse towards understanding on the one hand, and fear and its vain antidotes on the other. 
I believe I ``understand'', or at least \textbf{know} this impulse and its origin. (Or
perhaps one day I will discover the extent to which I was deluded).  
However, I know that I have not yet gotten to the bottom of the great enigma that is fear and vanity, 
and the insidious creativity blocks that result from them - 
and who knows if I will ever reach the conclusion of this mystery in the years I have
left\ldots

As I was writing 
R\'ecoltes et Semailles, 
two images emerged when attempting to represent these two aspects of the human journey:
the image of the \textbf{child} (aka the \textbf{laborer}), and that of the
\textbf{Boss}.
In the Walk which we are about to undertake, we will be dealing mostly with the ``child''.
He is also the one featuring in the subtitle ``\textbf{The Child and the Mother}''. 
The motivation behind this name 
will hopefully become clear over the course of the walk.

\marginpar{p. A4}
In the remainder of the reflection, however, the Boss will take the lead. 
He is living up to his name! It would be more accurate 
to speak of multiple Bosses of competing enterprises rather than of a \textbf{singular}
Boss.
But it is also true that all Bosses share the same essential features.
And once Bosses have been mentioned, it naturally follows that there will also be ``villains''.
In part I of the reflection (which is named ``Complacency and Renewal'' and will follow the present
introductory section, ``Prelude in four Parts"), I chiefly play the role of the ``villain''. In the following
three parts, it is mostly the ``others''. 
Each in turn!

That is to say, in addition to philosophical reflections and (unapologetic) ``confessions'', 
there will be ``vitriolic portraits'' (to use the expression of one of my
colleagues who found himself mistreated\ldots). 
Not to mention large-scale and well-oiled ``operations''.
Robert Jaulin\footnote{Robert Jaulin is an old friend of mine. From what I understand, his position within the establishment of the ethnology world (as a ``white wolf") mirrors the one I occupy within the mathematical ``high society''.} assured me (half-jokingly) that in 
R\'ecoltes et Semailles I was dressing the ``ethnology
of the mathematical world'' (or maybe the sociology, I don't quite remember).
It is flattering of course to learn
that one has been (unknowingly) doing scholarly things!
It is true that during the ``investigation'' segment of the reflection I mentioned in passing,
in the pages I was writing, a good chunk of the mathematical establishment,
as well as a number of my colleagues and friends of a more modest status.
Over the past few months, since I started sending preliminary versions of
R\'ecoltes et Semailles, 
this comment was ``brought up'' again. My testimony arrived like a rock landing in a pond. 
There were returns of every kind (except for boredom\ldots). 
Yet, almost every time, the response was far from what I had expected. 
There was also a lot of silence, which speaks volumes. 
Visibly, I had (and still have) a lot to learn about what happens in people's minds, among
my previous students and other colleagues - or should I say: I still have a lot to learn about ``the sociology of the mathematical milieu''!
To all those who contributed to the great sociological project of my old days, I would like
to express my sincere gratitude.

I was of course particularly touched by warm responses. There were also some rare
colleagues who opened up about a (previously unexpressed) feeling or sentiment of crisis, or a 
\marginpar{p. A5}
degradation, within the mathematical milieu to which they feel that they belong. 

Outside of this milieu, I would like to acknowledge 
Sylvie et Catherine Chevalley\footnote{Sylvie et Catherine Chevalley are the widow and
daughter, respectively, of Claude Chevalley, the colleague and friend to whom the central part of 
R\'ecoltes et Semailles is devoted (ReS III, the key of the Yin and the Yang).
I write about him and the role that he played in my
journey at several points in the reflection.} Robert Jaulin, St\'ephane Deligeorge, Christian Bourgois for being among the very first people to respond warmly, or even affectionately, to my testimony. 
If R\'ecoltes et Semailles achieves a wider diffusion than that of the initial 
printing (which was addressed to a very limited social circle), it will be mostly thanks to
them - thanks, mostly, to their communicative 
conviction that what I strived to grasp and verbalize had to be said, and that my message could find an audience wider than just my colleagues (who are often sullen,
if not belligerent, and strictly opposed to questioning their position\ldots).
Case in point, Christian Bourgois unhesitatingly decided to run the risk of publishing the unpublishable, and St\'ephane Deligeorge to place my 
indigestible testimony alongside works of Newton, Cuivier, and Arago (I could not have asked for
better company!).
To each of them, for their repeated expressions of sympathy and trust offered at an
especially ``sensitive" moment, I hereby gladly express all my gratitude. 

And so here we are, about to begin a Walk through my life's work, which shall serve as a primer to
a journey through my life. 
A long journey, to be sure, spanning over a thousand densely packed pages. I spent a lifetime undergoing this journey, without it ever 
coming to an end, and it took me more than a year to rediscover it, 
one page at a time.
Words were sometimes difficult to come by, as I sought to convey an experience which continues to evade my comprehension - just as ripe and thick grapes stacked in a press sometimes seem to be evading the force applied upon them\ldots
But even at times when the words seemed to come flooding, nothing was left to chance. 
Each word was gauged in the process, or after the fact, so as to obtain a carefully calibrated end result. As such, this reflection/testimony/journey is not meant to be read hastily, in a day or in a
month, by a reader who is in a rush to reach the final word. There is \textbf{no} 
\marginpar{p. A6}
``final word'', no
``conclusion'' in 
R\'ecoltes et Semailles, no more than there are any such things in my life or in yours. 
What you will find instead is a wine, aged over the course of a lifetime in the depths of my being.
The last glass which you will be drinking will be no better nor worse than the first or the
hundredth. 
They are all ``the same'', just as they are all different. And if the first glass is spoiled,
so is the rest of the barrel; better to drink fresh water (if such can be found),
than to drink bad wine. 

But a good wine ought not to be drunk in haste, nor 
expeditiously. 

%\end{document}
